---
title: "Relationship between the size of consonant inventory and complexity of syllable structure"
subtitle: "An exploration of WALS database using Bayesian regression models"
css: styles.css
---

# 1. Introduction

## 1.1 WALS database and relevant linguistic features
This project is based on the chapter 12 of the World Atlas of Language Structures (WALS) database (Maddieson, 2013), which contains many features describing aspects of cross-linguistic diversity. It aims to further explore a topic mentioned in that chapter: the relationship between the size of consonant inventory and complexity of the syllable structure.

<https://wals.info/chapter/12> presents the definition of consonant inventory and that of syllable structure complexity. The former is quite straightforward: the number of consonants in a given language; as for the syllable structure complexity, it is divided into three categories: simple (CV, i.e., a consonant preceding a vowel), moderately complex(CCVC, with some limitations on the second consonant), and complex (several consonants before and after a vowel, which is expressed as (C)(C)(C)V(C)(C)(C)(C)).

(Maddieson, 2013) has mentioned a "significant though not strong" positive correlation between the size of consonant inventory and the syllable structure data: "the correlation coefficient is 0.203 and the statistical significance level is very high, p<0.0001."

## 1.2 Research questions
The existing results raise two important research questions:

First, the conclusion of positive correlation has been drawn from frequentist statistics; however, a single small p-value may sometimes lead to an unreliable conclusion, thus it is reasonable to use Bayesian statistics to check if such a correlation really exists. The advantage of using a Bayesian regression model is that, instead of providing a fixed correlation coefficient, it allows us to get a posterior distribution of the correlation parameters, which makes it possible to assess the credibility of the relationship between these variables.

The second question has been pointed out by the author at the end of this chapter as a topic for further exploration: "Analysis of the patterns within and between different linguistic areas and language families would be required to decide whether the association should be attributed to accidents of survival and spread of particular languages, or can be proposed as reflecting a design feature of language viewed as a whole." As indicated by the author, the pattern of such a positive correlation comes from the areal overlaps between small consonant inventories and simple syllable structure, and large consonant inventories and complex syllable structure. For this reason, when fitting a Bayesian regression model, it is necessary to use geographic areas as an interaction term: even if an overall positive correlation does exist, it may not be a universal pattern if the relationship actually differs among different areas. 

This research aims to figure out whether or not such positive correlation exists under Bayesian regression, and if it does exist, whether or not it is really a universal pattern regardless of geographic areas.


# 2. Data acquisition and wrangling
## 2.1 The data from WALS
The data from WALS can be easily accessed by the R package ritwals (Kauhanen, 2024).
```{r}
library(ritwals)
Syllable_Structure <- get_feature('12A', data = WALS)
```
However, the data available in WALS are not sufficient for the implementation of this project. The feature of consonant inventories in WALS only takes five categorical values: small, moderately small, average, moderately large and large, with their corresponding intervals. Without an exact number of consonants, this feature is too simplified to be used in modelling the relationship between the size of consonant inventory and the complexity of syllable structure. For this reason, in this research only the syllable structure feature is taken from the chapter 12 of WALS (Maddieson, 2013), and another database is required to provide the number of consonants in each language.

## 2.2 The data from PHOIBLE
The file `phoible.csv`, provided by the database PHOIBLE and available at <https://github.com/phoible/dev/tree/master/raw-data/PH>, contains phonological inventory data (phonemes, allophones and tones) drawn from journal articles, theses, and published grammars. As it contains all the phonemes in a certain language and the `SegmentClass` column indicates whether the phoneme is consonant or vowel, with some data wrangling process it is easy to get the number of consonants of each language in this file.
```{r}
library(conflicted) # Use the conflicted package to force all conflicts to become errors
library(tidyverse)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::summarise)

phoible <- read.csv('./phoible.csv')
```

In both the data from WALS and that from PHOIBLE, each language has a Glottocode that distinguishes it from other languages. Therefore, the two data sets should be merged according to the same language name and the same Glottocode for this research. However, `phoible.csv` has much richer data as it may include different dialects of the same language: the dialects of a certain language are under the same Glottocode, and are specified in the column `SpecificDialect`. As the WALS data only contain information about the standard version of each language, the dialect variations in PHOIBLE data should be discarded in this case.
```{r}
phoible <- phoible[is.na(phoible$SpecificDialect),]
```

The `distinct()` function from `dplyr` helps filter out unique rows from the phoible data frame. The uniqueness is determined by the combined values of three columns: `Glottocode`, `LanguageName`, and `Phoneme`.
```{r}
phoible <- phoible |>
  distinct(Glottocode, LanguageName, Phoneme, .keep_all = TRUE)
```

After ensuring that under the same `Glottocode` and `LanguageName` each phoneme is unique, we can count the number of consonants in each language.
```{r}
cons_count <- phoible |>
  dplyr::filter(SegmentClass == 'consonant') |>
  group_by(Glottocode, LanguageName, SegmentClass) |>
  summarise(count = n(), .groups = 'drop') |>
  mutate(consonant_count = count)

cons_count <- cons_count[c('Glottocode', 'LanguageName', 'consonant_count')]
``` 

## 2.3 Joining the two datasets
With the number of consonants in each language that is uniquely specified by its `LanguageName` and `Glottocode`, we can join the two data frames to get the data required by this research topic. The `inner_join()` function from `dplyr` joins the two data frames and only returns those rows that have matching keys in both data frames. In this way, for each language in the processed data, the data contain both the complexity of syllable structure and the exact number of consonants.
```{r}
# Convert all the language names into lowercase.
Syllable_Structure$language <- tolower(Syllable_Structure$language)
cons_count$LanguageName <- tolower(cons_count$LanguageName)
languages_data <- inner_join(Syllable_Structure, cons_count, by = c("language" = "LanguageName", "glottocode" = "Glottocode")) |>
  drop_na()
languages_data <- languages_data |>
  mutate(complexity = value) |>
  dplyr::select(-value) # use a more recognizable name

languages_data$complexity <- factor(languages_data$complexity, levels = c("Simple", "Moderately complex", "Complex"), ordered = TRUE) # set a more reasonable order of values
```

The `languages_data` object is now ready-to-use for further analysis.


# 3. Basic analysis of the prepared data
In this section, `ggplot` package is used to make some basic plots to explore the information in the data. With the `plotly` package, the plots are set interactive.

## 3.1 Syllable structure complexity and average consonants
```{r}
languages_data |> group_by(complexity) |>
  summarise(average_consonants = mean(consonant_count),
            number_of_languages = n())
```

(Maddieson, 2013) has mentioned that, based on their data, the languages with simple, moderately complex and complex structure have an average of 19.1, 22.0 and 25.8 consonants, respectively. These numbers do not exactly correspond to the data in this research, because the latter is the intersection of two data sets and thus does not contain exactly the same languages as WALS database. However, the trend of the number of consonants varying with syllable complexity is consistent: languages with more complex syllable structure tend to have a larger inventory of consonants. Section 4 will use Bayesian regression model to investigate whether such trend is a universal property for all the languages around the world.

```{r}
library(ggplot2)
library(plotly)
violin_average <- languages_data |>
  ggplot(mapping=aes(x=complexity, y=consonant_count,
                     fill=complexity)) + 
  geom_violin() +
  scale_fill_brewer(palette = "YlOrRd") +
  geom_jitter(alpha = 0.5,
              mapping = aes(text =  paste("Language:", language,
                         "<br>Complexity:", complexity,
                         "<br>Consonant Count:", consonant_count)))

ggplotly(violin_average, tooltip = "text")
```

The violin plot shows the distribution of consonant counts for language counts categorised by syllable structure complexity.
Although the most concentrated range of consonant counts increases progressively across the three syllable structure categories, the majority of languages in each category have relatively moderate consonant counts, and it is rare to have exceptionally high counts. This can be supported by (Jakobson et al., 1952): languages avoid overly complex systems to ensure the perceptual distinctiveness and economy.

## 3.2 Syllable structure complexity and macroarea
In the data for this research there are 6 macroareas in total: Africa, Australia, Eurasia, North America, Papunesia and South America. Bar plot can be used to demonstrate the proportion of each syllable structure complexity categories in different macroareas.
```{r}
categories_in_macroareas <- languages_data |> ggplot(mapping=aes(x=macroarea, fill=complexity)) +
  geom_bar(position='dodge') +
  scale_fill_brewer(palette = "YlOrRd") + 
  labs(title = "Syllable Structure Complexity Distribution Across Macroareas")
ggplotly(categories_in_macroareas)
```

The plot above illustrates that the proportions of different syllable structure complexity categories differ among macroareas. From another perspective, we can also plot the macroarea distribution for each syllable structure complexity:
```{r}
macroareas_in_categories <- languages_data |> ggplot(mapping=aes(x=complexity, fill=macroarea)) +
  geom_bar(position='dodge') +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Macroarea Distribution within Each Syllable Structure Complexity Category")
ggplotly(macroareas_in_categories)
```

The two plots above indicate that, for some macroareas, there could be certain syllable structure complexity that is very rare or even absent. For example, according to the data of this research, in Australia and North America there is no language with a simple syllable structure, and in Eurasia there is only one language with this structure. Such sparse data determine the limitation of this research: the unbalanced distributions across complexity categories may make the regression results less reliable.


# 4. Bayesian Regression
## 4.1 Model selection
With the R package `brms` (Bürkner, 2018), the data in this research are analysed by fitting a Bayesian regression model. More specifically, here it is ideal to choose an adjacent categorical model, because the complexity of syllable structure is an ordered variable (from "simple" to "moderately complex" to "complex") but does not follow a strictly uniform progression: for example, based on the definition of (Maddieson, 2013), compared to the difference between "simple" and "moderately complex" syllable structure, the transition from "moderately complex" to "complex" involves more freedom in the combination of consonants and vowels. As discussed in the STAT 504 course from Pennsylvania State University (n.d.), adjacent category models are ideal for situations where ordered categories do not meet proportional odd assumptions. This flexibility makes this model a suitable choice in our context.

Section 4.2 will first simply investigate the relationship between the outcome and the size of consonant inventory, which is treated as a continuous numeric predictor.

```{r}
library(brms)
conflicts_prefer(brms::ar)
```

## 4.2 Fit an adjacent categorical model
As the syllable complexity is an ordered categorical variable, an adjacent categorical model is fitted to check the relationship between this outcome and the size of consonant inventory. 
```{r}
# syllable structure and consonants
ss_and_consonants <- brm(
  formula = complexity ~ consonant_count,
  data = languages_data,
  family = acat(link = "logit"),
  cores = 4,
  seed = 1229,
  file = "cache/ss_and_consonants"
  )
summary(ss_and_consonants)
```

The intercepts represent the first pair of adjacent categories (simple vs. moderately complex) and the second pair of adjacent categories (moderately complex vs. complex). These two estimated values show that the "moderately complex" category has the highest baseline probability.

According to the model, both the log odds of transitioning from "simple" to "moderately complex" syllable structure and that of transitioning from "moderately complex" to "complex" increase by 0.04 for every additional consonant in the inventory, with a 95% credible interval ranging from 0.02 to 0.06. Although the 95% credible interval does not contain zero, which allows us to interpret the evidence as reliable (Kruschke et al., 2012), the effect size is relatively small, suggesting that the consonant inventory size has a weak influence on syllable complexity.

The function `conditional_effects()` from the `brms` package is used to visualise the relationship between the outcome and the predictor. The argument `categorical=TRUE` is included to generate marginal probability plots for each category of syllable structure complexity, allowing us to visualise how the predicted probabilities of each category change with the consonant inventory size.

```{r}
conditional_effects(ss_and_consonants, categorical = TRUE)
```

As the plot indicates, when the consonant inventory size increases, the probability of a complex structure increases and the probability of simple or moderately complex structure decreases.

When the consonant size is relatively small (less than 40), the most probable structure is a moderately complex structure. This is because of the proportions in the data: it is mentioned above that the "moderately complex" category has the highest baseline probability. It is worth noting that the probability of a simple structure decreases faster than the probability of a moderately complex structure. This is consistent with the positive log odds of transitioning from "simple" to "moderately complex" syllable structure, which means that with the increasing number of consonants, languages are more likely to have a moderately complex structure rather than a single structure.

When the consonant inventory size goes larger(more than 40), a complex structure is the most probable syllable structure. However, as indicated by the violin plot in section 3.1, most languages do not have so large an inventory size. 

The results in this section has basically verified the conclusion of (Maddieson, 2013), i.e., there is a significant though not strong correlation between these two variables. 

## 4.3 Macroarea as interactive predictor
Section 4.2 has verified the significant though not strong correlation between syllable structure complexity and consonant inventory size. In this section, `macroarea` category will be included as an interactive predictor to figure out whether this correlation is a universal property around the world. 

In this analysis, the `macroarea` predictor will be modelled by a dummy variable approach, as the adjacent category model does not allow the removal of the baseline category. By using dummy variables, we are able to estimate the effects for each macroarea independently, making the results more straightforward.
```{r}
languages_data <- languages_data |>
  mutate(
    macroarea_Africa = ifelse(macroarea == "Africa", 1, 0),
    macroarea_Australia = ifelse(macroarea == "Australia", 1, 0),
    macroarea_Eurasia = ifelse(macroarea == "Eurasia", 1, 0),
    macroarea_NorthAmerica = ifelse(macroarea == "North America", 1, 0),
    macroarea_Papunesia = ifelse(macroarea == "Papunesia", 1, 0),
    macroarea_SouthAmerica = ifelse(macroarea == "South America", 1, 0)
    )

ss_and_consonants_with_macroarea <- brm(
  formula = complexity ~ consonant_count:macroarea_Africa +
                         consonant_count:macroarea_Australia +
                         consonant_count:macroarea_Eurasia +
                         consonant_count:macroarea_NorthAmerica +
                         consonant_count:macroarea_Papunesia +
                         consonant_count:macroarea_SouthAmerica,
  data = languages_data,
  family = acat(link = "logit"),
  cores = 4,
  seed = 1229,
  file = "cache/ss_and_consonants_with_macroarea"
  )
summary(ss_and_consonants_with_macroarea)
```

To summarise, the relationship between the syllable structure complexity and the consonant inventory size differs among different macroareas:

- **Significant positive effects**:
In Australia, Eurasia and North America, `consonant_count` shows a relatively positive effect on syllable complexity, with a 95% credible interval that excludes 0:
Australia: Estimate = 0.06; 95% CI: 0.02 to 0.08;
Eurasia: Estimate = 0.06; 95% CI: 0.04 to 0.08;
North America: Estimate = 0.06; 95% CI: 0.03 to 0.09.
This indicates that, in these three macroareas, larger consonant inventory increases the log odds of transitioning to more complex syllable structures.

- **No significant effects**:
In contrast, in Africa, Papunesia and South America, the effects are not significant, with 95% credible intervals including 0:
Africa: Estimate = -0.00; 95% CI: -0.02 to 0.02;
Papunesia: Estimate = 0.01; 95% CI: -0.04 to 0.05;
More specifically, the estimate for South America is negative (-0.03); however, this effect is not so significant either, as the 95% CI is from -0.08 to 0.02. In conclusion, in these macroareas, the consonant inventory size has no meaningful impact on syllable structure complexity.


# 5. Conclusion
By using an adjacent category Bayesian regression model, this research has verified the conclusion of (Maddieson, 2013) based on frequentist statistics: there does exist a positive relationship between the syllable structure complexity and the size of consonant inventory, although the relationship is not strong. It should be noted, however, that this finding reflects an overall pattern when not distinguishing between different geographic regions: with the macroarea as an interactive predictor, further explorations have demonstrated that this positive relationship is only found in Australia, Eurasia and North America, but it is not significant in Africa, Papunesia and South America.

Given that the positive relationship does not hold across all macroareas, and is relatively weak even if it exists in some regions, it suggests that consonant inventory size is not likely to be a decisive factor of the syllable structure complexity. Overall, most languages do not maintain an exceptionally large consonant inventory, and it is most common to have a moderately complex syllable structure, followed by a complex structure. A simple structure is the rarest among three categories and tend to be more common in certain specific regions, like Africa and Papunesia. Such preference reflects the tendency to reduce the total number of phonemes by expanding the expressions making more complex combinations of vowels and consonants.


# 6. References

[1] Bürkner, P.-C. (2018). Advanced Bayesian multilevel modeling with the R package brms. The R Journal, 10(1), 395–411. https://doi.org/10.32614/RJ-2018-017

[2] Kauhanen H (2024). _ritwals: R Interface to WALS_. R package version
  0.0.0.9102, commit af5d57eaf4a88527b167a465841f7b96225738b0,
  <https://github.com/hkauhanen/ritwals>.
  
[3] Kruschke, J. K., Aguinis, H., and Joo, H. (2012). The time has come: Bayesian methods for data analysis in the organizational sciences. Org. Res. Methods 15, 722–752. doi:10.1177/1094428112457829

[4] Maddieson, I. (2013). Syllable structure. In M. S. Dryer & M. Haspelmath (Eds.), WALS Online (v2020.4) [Data set]. Zenodo. https://doi.org/10.5281/zenodo.13950591 (Available online at http://wals.info/chapter/12. Accessed on 2024-11-02.)

[5] PHOIBLE. (n.d.). PH (raw-data) [Data set]. GitHub. Retrieved November 2, 2024, from https://github.com/phoible/dev/tree/master/raw-data/PH

[6] Pennsylvania State University. (n.d.). 8.3: Adjacent categories logit models. STAT 504: Analysis of discrete data. Retrieved December 5, 2024, from https://online.stat.psu.edu/stat504/lesson/8/8.3
